Nail: A practical interface generator for data formats

### 2014-11-14 paper/login/eval.tex
Login first rough draft

\input{fig-ethernet}

In our evaluation of Nail, we answer four questions:

\begin{itemize}

\item Can Nail grammars support real-world data formats, and
      are Nail's techniques critical to handling these formats?

\item How much programmer effort is required to build an
      application that uses Nail for data input and output?

\item Does using Nail for handling input and output improve
      application security?

\item Does Nail achieve acceptable performance?

\end{itemize}

\subsection{Data formats}
\label{s:eval-formats}

To answer the first question, we used Nail to implement grammars
for seven protocols with a range of challenging features.
Figure~\ref{fig:eval-protocols} summarizes these protocols, the lines
of code for their Nail grammars, and the challenging features that make
the protocols difficult to parse with state-of-the-art parser generators.
We find that despite the challenging aspects of these protocols, Nail is
able to capture the protocols, by relying on its novel features: dependent
fields, streams, and transforms.  In contrast, state-of-the-art parser
generators would be unable to fully handle 5 out of the 7 data formats.
In the rest of this subsection, we describe the DNS and Zip grammars in
more detail, focusing on how Nail's features enable us to support these
formats.

@@

In Section~\ref{s:design}, we introduced Nail's syntax with a grammar
for DNS packets, shown in Figure~\ref{fig:dns-xform}.
The grammar corresponds almost directly to the diagrams in RFC 1035,
which defines DNS~\cite[\S4]{RFC:1035}.  Each DNS packet consists of a
header, a set of question records, and a set of answer records. Domain
names in both queries and answers are encoded as a sequence of labels,
terminated by a zero byte.  Labels are Pascal-style strings, consisting
of a length field followed by that many bytes comprising the label.

One challenging aspect of DNS packets lies in the count fields (\cc{qc},
\cc{ac}, \cc{ns}, and \cc{ar}), which represent the number of questions
or answers in another part of the packet.  Nail's \cc{n_of} combinator
handles this situation easily, which would have been difficult to handle
for other parsers.

Another challenging aspect of DNS is label
compression~\cite[\S4.1.4]{RFC:1035}.  Label compression is used to reduce
the size overhead of including each domain name multiple times in a DNS
reply (once in the question section, and at least once in the response
section).  If a domain name suffix is repeated, instead of repeating that
suffix, the DNS packet may write a two-bit marker sequence followed by
a 14-bit offset into the packet, indicating the position of where that
suffix was previously encoded.

Handling label compression in existing tools, such as Bison or Hammer,
would at best be very awkward, because some ad-hoc trick would have to
be used to re-position the parser's input stream.  Keeping track of the
position of all recognized labels would not be enough, as the offset field
may refer to any byte within the packet, not just the beginning of labels.
For this reason, the DNS server used as the example for Hammer does not
support compression.

\begin{figure}
\smaller[0.5]
\input{code/dns-xform}
\caption{Signatures of stream transform functions for handling DNS label compression.}
\label{fig:dns-xform}
\end{figure}

In contrast, Nail is able to handle label compression, by using a stream
transform; the signatures of the two transform functions are shown
in Figure~\ref{fig:dns-xform}.  When parsing a packet, this transform
decompresses the DNS label stream by following the offset pointers.
When generating a packet, this transform receives the current suffix as
an input, and scans the packet so far for previous occurrences, which
implements compression.

\paragraph{ZIP files.}

An especially tricky data format is the ZIP compressed archive
format~\cite{pkzip}.  ZIP files are normally parsed end-to-beginning. At
the end of each ZIP file is an \emph{end-of-directory header}. This header
contains a variable-length comment, so it has to be located by scanning
backwards from the end of the file until a magic number and a valid
length field is found. Many ZIP implementations disagree on how to find
this header in confusing situations, such as when the comment contains
the magic number~\cite{wolf:berlinsides-zip}.

 This end-of-directory
header contains the offset and size of the \emph{ZIP directory}, which
is an array of \emph{directory entry headers}, one for every file
in the archive.  Each entry stores file metadata, such as file name,
compressed and uncompressed size, and a checksum, in addition to the
offset of a \emph{local file header}. The local file header duplicates
most information from the directory entry header. The compressed file
contents follow the header immediately.

\input{fig-zip}

Duplicating information made sense when ZIP files were stored on floppy
disks with slow seek times and high fault rates, and memory constraints
made it impossible to keep the ZIP directory in memory or the archive
was split across multiple disks.  However, care must be taken that the
metadata is consistent. For example, vulnerabilities could occur if
the length in the central directory is used to allocate memory and the
length in the local directory is used to extract without checking that
they are equal first, as was the case in the Python ZIP
library~\cite{python-bug:20078}.
\begin{comment}
\begin{figure}[tb]
\smaller[0.5]
\input{code/zip-eod-xform}
\caption{Signatures of stream transform functions for handling the
end-to-beginning structure of ZIP files.}
\label{fig:zip-eod-xform}
\end{figure}
\end{comment}
Figure~\ref{fig:zip-extract} shows an abbreviated version of our ZIP
file grammar.  The ZIP grammar is a good example of how transformations
capture complicated syntax in a real-world file format; existing parser
languages cannot handle a file format of this complexity.

The \cc{zip\_file} grammar first splits the entire file stream into two
streams based on the \cc{zip\_end\_of\_directory} transform on line 2.
%(whose two C function signatures are shown in Figure~\ref{fig:zip-eod-xform}). 
\cc{zip\_end\_of\_directory_parse}
finds the end-of-directory header as described above, by scanning the
file backwards, and splits the file into two streams, one containing
the end-of-directory header and one containing the file contents. The
\cc{end\_of\_directory} rule is then applied to the header stream in
line 4. All offsets in the ZIP file refer to the beginning of the
file, so the stream \cc{\$file} which contains the file contents
without the header is passed as an argument to all parsers from
hereon.

The directory header contains the offset and size of the ZIP
directory (lines 9 and 10). The \cc{offset} and \cc{size}
transformations extract a stream containing just the directory from
the file contents. This stream is then parsed as an array of directory
entries in line 17.
Each directory entry in turn points to a local file header, which is
similarly extracted and parsed with the \cc{file} rule.

The \cc{file} rule starting at line 41, describing a ZIP file entry,
takes dependent field
parameters containing file metadata information from the directory
header. However, this same information is duplicated in the file
entry, so the grammar uses the Nail-supplied \cc{u16\_depend}
transform to check whether the two values are equal. Unlike most other
transforms, \cc{u16\_depend} does not consume or produce strings; it
only checks that two dependent fields are equal when parsing, and
assigns the value of the second field to the first when generating.
This ensures that the programmer does not have to worry about
inconsistencies when handling the internal representation of a ZIP
file.

Immediately following the file entry is the compressed data.  Because
most compression algorithms operate on unbounded streams of data, Nail
decompresses data in two steps.  First, it isolates the compressed
data from the rest of the stream by using the \cc{size} transform,
which operates on the current stream, meaning it will consume data
starting at the current position of the parser in the input.  Second,
Nail invokes a custom \cc{zip\_compression} transform that implements
the appropriate compression and decompression functions based on the
specified compression method. These functions are otherwise oblivious
to the layout or metadata of the file.

\subsection{Programmer effort}
\label{s:eval-effort}

%code size unzip 6.0
%2821 lines

To evaluate how much programmer effort is required to build an
application that uses Nail, we implemented two applications---a
DNS server, and an
\cc{unzip} program---based on the above grammars, and compared code
size with comparable applications that process data manually, using
\cc{sloccount}~\cite{sloccount}.  We also compare the code size of our
DNS server to a DNS server written using the Hammer parsing framework,
although it does not fully support DNS (e.g., it lacks label compression,
among other things).  Figure~\ref{fig:effort} summarizes the results.

\begin{figure}[tb]
\centering
\smaller[0.5]
\begin{tabular}{@{~}lrr@{~}l@{~}}
\toprule
\textbf{Application}
  & \textbf{LoC w/ Nail}
  & \multicolumn{2}{c}{\textbf{LoC w/o Nail}} \\
\midrule
DNS server
  & 295
  & 683
  & (Hammer parser) \\

@@

% It is hard to compare the programming effort required to implement
% our DNS server to that of a real world DNS server, since we implement
% less functionality: we do not send the additional glue records and
% we did not implement any configuration options.  However, the closest
% in functionality and intent is Dan Bernstein's djbdns~\cite{djbdns},
% which aims to be a minimalist, highly secure, authoritative-only DNS
% server. The latest release of djbdns, including various support tools,
% is about 10,000 lines of C\@.  \XXX[nz][so what does this mean?]

\paragraph{ZIP.}

We implemented a ZIP file extractor in 50 lines of C code, together with
92 lines of Nail grammar and 78 lines of C code implementing two stream
transforms (one for the DEFLATE compression algorithm with the help of
the zlib library, and one for finding the end-of-directory header).

Because more recent versions of ZIP have added more features, such
as large file support and encryption, the closest existing tool
in functionality is the historic version 5.4 of the Info-Zip unzip
utility~\cite{infozip} that is shipped with most Linux distributions.
The entire unzip distribution is about 46,000 lines of code, which is
mostly optimized implementations of various compression algorithms and
other configuration and portability code.  However, unzip isolates the
equivalent of our Nail tool in the file \cc{extract.c}, which parses the
ZIP metadata and calls various decompression routines in other files. This
file measures over 1,600 lines of C, which suggests that Nail is highly
effective at reducing manual input parsing code, even for the complex
ZIP file format.

\subsection{Security}

We use a twofold approach to evaluate the security of applications
implemented with Nail. First, we analyze a list of CVE's related to the
ZIP file format and argue how our ZIP tools based on Nail are immune
against those vulnerability classes. Second, we present the results of
fuzz-testing our DNS server.

\paragraph{ZIP analysis.}

In \S\ref{s:motivation}, we found 15 input handling vulnerabilities
related to ZIP files.

11 of these vulnerabilities involved memory corruption during input handling.
Because Nail's generated code checks offsets before reading and does not
expose any untrusted pointers to the application, it is immune against
memory corruption attacks by design.

Nail also protects against parsing inconsistency vulnerabilities like the four others we studied. First,
Nail grammars explicitly encode duplicated information such as the redundant length fields in ZIP
that caused a vulnerability in the Python ZIP library. Second, the other three vulnerabilities exist
because multiple implementations of the same protocol disagree on some inputs. Hand-written protocol
parsers are not very reusable, as they build application specific data structures and are tightly
coupled to the rest of the code. Nail grammars, however, can be re-used between applications,
avoiding protocol misunderstandings.

\paragraph{DNS fuzzing.}

To provide additional assurance that Nail parsers are free of memory
corruption attacks, we ran the DNS fuzzer provided with the Metasploit
framework~\cite{mspframework} on our DNS server, which sent randomly
corrupted DNS queries to our server for 4 hours, during which it did
not crash or trigger the stack or heap corruption detector.

\subsection{Performance}

To evaluate whether Nail-based parsers are compatible with good
performance, we compare the performance of our DNS server to that of ISC
BIND 9 release 9.9.5~\cite{bind9}, a mature and widely used DNS server.
We simulate a load resembling that of an authoritative name server.
First, we generate domain names consisting of one or two labels randomly
selected from an English dictionary, and one label that is one of
three popular top-level domains (\cc{com}, \cc{net}, and \cc{org}).
Second, we randomly selected 90\% of these domains and created a zone
file that mapped these domain names to \cc{127.0.0.1}.  Finally, we used
the \cc{queryperf} tool provided with BIND to query each domain between
zero and three times, using a DNS server running on the local machine.
We used a single core of an Intel i7-3610QM system with 12GB of RAM\@.
The benchmark tool kept at most 20 queries outstanding at once, and
was configured to repeat the same randomized sequence of queries for
one minute.  We repeated each test seven times with 50,000 domain names,
restarting each daemon in between; we also repeated the tests with 1
million domain names, and found similar results.  We also performed one
initial dry run to warm the file system cache for the zone file.

The results are shown in Figure~\ref{fig:perf-dns}, and demonstrate that
our Nail-based DNS server can achieve higher performance and lower latency
than BIND\@.  Although BIND is a more sophisticated DNS server,
and implements many features that are not present in our Nail-based DNS
server and allow it to be used in more complicated configurations, 
we believe our results demonstrate that Nail's parsers are not
a barrier to achieving good performance.

\input{fig-perf-dns}

>>>>

% In our evaluation of Nail, we answer four questions:

% \begin{itemize}

% \item Can Nail grammars support real-world data formats, and
%       are Nail's techniques critical to handling these formats?

% \item How much programmer effort is required to build an
%       application that uses Nail for data input and output?

% \item Does using Nail for handling input and output improve
%       application security?

% \item Does Nail achieve acceptable performance?

% \end{itemize}

\subsection{Real-World Formats}
\label{s:eval-formats}

First, we ask whether Nail can handle real-world data formats and whether its techniques are
critical to that. We used Nail to implement grammars
for seven protocols with a range of challenging features.
Figure~\ref{fig:eval-protocols} summarizes our results.
We find that despite the challenging aspects of these protocols, Nail is
able to capture the protocols, by relying on its novel features: dependent
fields, streams, and transforms.  In contrast, state-of-the-art parser
generators would be unable to fully handle 5 out of the 7 data formats.
In the rest of this subsection, we describe the DNS and Zip grammars in
more detail, focusing on how Nail's features enable us to support these
formats.

@@

\paragraph{DNS} In Section~\ref{s:design}, we introduced Nail's syntax with a grammar
for DNS packets, shown in Figure~\ref{fig:dns-full}.
The grammar corresponds almost directly to the diagrams in RFC 1035,
which defines DNS~\cite[\S4]{RFC:1035}. Nail's dependent fields handle DNS's count fields, and
transformations represent label compression. Both of these features are awkward at best to handle
with existing tools.

\paragraph{ZIP} An especially tricky data format is the ZIP compressed archive format~\cite{pkzip}.
ZIP files are normally parsed end-to-beginning. At the end of each ZIP file is an
\emph{end-of-directory header}. This header contains a variable-length comment, so it has to be
located by scanning backwards from the end of the file until a magic number and a valid length field
is found. Many ZIP implementations disagree on the exact semantics of this, such as when the comment
contains the magic number~\cite{wolf:berlinsides-zip}.

This header  contains points to the \emph{ZIP directory}, contains an entry every file
in the archive.  Each entry stores file metadata in addition to the
offset of a \emph{local file header}. The local file header duplicates
most information from the directory entry header and is followed immediately by the compressed
archive entry.
Duplicating information made sense when ZIP files were stored on floppy
disks with slow seek times and high fault rates, and memory constraints
made it impossible to keep the ZIP directory in memory.

 However, if attackers craft inconsistent files, hand written parsers fail in interesting ways, e.g. 
the length in the central directory is used to allocate memory and the
length in the local directory is used to extract without checking that
they are equal first, as was the case in the Python ZIP
library~\cite{python-bug:20078}.

Nail avoids these issues by capturing these redundancies with dependent fields and decompressing
files transparently with transformations, and the programmer
is presented with a data structure containing the archive contents. To handle formats based on ZIP,
such as office documents, our ZIP grammar could be extended to apply additional grammars to the
files within an archive.

\subsection{Applications}
\label{s:eval-effort}

%code size unzip 6.0
%2821 lines
We implemented two applications---a
DNS server and an
\cc{unzip} program---based on the above grammars, and will compare the effort involved and the
resulting security to similar applications with hand-written parsers and with other parser generators.
We will use lines of code as a proxy for programmer effort. To evaluate security, we will argue how
our design avoids classes of vulnerabilities and fuzz-test one of our applications.
\begin{figure}[tb]
\centering
\smaller[0.5]
\begin{tabular}{lrr@{~}l}
\toprule
\textbf{Application}
  & \textbf{LoC w/ Nail}
  & \multicolumn{2}{c}{\textbf{LoC w/o Nail}} \\
\midrule
DNS server
  & 295
  & 683
  & (Hammer parser) \\

@@

We ran the DNS fuzzer provided with the Metasploit
framework~\cite{mspframework} on our DNS server, which sent randomly
corrupted DNS queries to our server for 4 hours, during which it did
not crash or trigger the stack or heap corruption detector.

\paragraph{Performance}

To evaluate whether Nail-based parsers are compatible with good
performance, we compare the performance of our DNS server to that of ISC
BIND 9 release 9.9.5~\cite{bind9}, a mature and widely used DNS server.
We simulate a load resembling that of an authoritative name server, generating a random zone file
and a random sequence of queries, with 10\% non-existent domains. We repeated this sequence of
queries for one minute against both DNS servers. The results are shown in Figure~\ref{fig:perf-dns}, and demonstrate that
our Nail-based DNS server can achieve higher performance and lower latency
than BIND\@.  Although BIND is a more sophisticated DNS server,
and implements many features that are not present in our Nail-based DNS
server and that allow it to be used in more complicated configurations, 
we believe our results demonstrate that Nail's parsers are not
a barrier to achieving good performance.
\paragraph{ZIP.}

We implemented a ZIP file extractor in 50 lines of C code, together with
92 lines of Nail grammar and 78 lines of C code implementing two stream
transforms (one for the DEFLATE compression algorithm with the help of
the zlib library, and one for finding the end-of-directory header).
 The \cc{unzip} utility contains a file \cc{extract.c}, which parses
ZIP metadata and calls various decompression routines in other files. This
file measures over 1,600 lines of C, which suggests that Nail is highly
effective at reducing manual input parsing code, even for the complex
ZIP file format.

In our full paper~\cite{bangert:nail-osdi14}, we present a study of 15 ZIP parsing bugs.
11 of these vulnerabilities involved memory corruption during input handling, which generated code is immune to
memory corruption attacks by design.
More interestingly, Nail also protects against parsing inconsistency vulnerabilities like the four others we studied.
Nail grammars explicitly encode duplicated information such as the redundant length fields in ZIP
that caused a vulnerability in the Python ZIP library. The other three vulnerabilities exist
because multiple implementations of the same protocol disagree on some inputs. Hand-written protocol
parsers are not very reusable, as they build application-specific data structures and are tightly
coupled to the rest of the code. Nail grammars, however, can be re-used between applications,
avoiding protocol misunderstandings.

### 2014-11-14 paper/login/impl.tex
Login first rough draft

\paragraph{Defense-in-Depth.}

Security exploits often rely on raw inputs being present in memory~\cite{shotgun-parser}, for example to include
shell-code or crafted stack frames for ROP~\cite{phrack58:4-nergal}-attacks in padding fields or the
application executing a controlled sequence of heap allocations and de-allocations to place
specific data at predictable addresses~\cite{jp-advanced, vudo-malloc}. Because the rest of the
application or even Nail's generated code may contain memory corruption bugs, Nail
carefully handles memory allocations as defense-in-depth to make exploiting such vulnerabilities harder.

When parsing input, Nail uses two separate memory arenas. These arenas allocate memory from the
system allocator in large, fixed size blocks. Allocations are handled linearly and all data in the
arena is zeroed and freed at the same time. Nail uses one arena for data used only during parsing,
including dependent fields and temporary streams and is released before the Nail's parser returns.
The other arena is used to allocate the internal data type returned and is freed by the application
once it is done processing an input.

Furthermore, the internal representation does not include any references to the input stream, which
can therefore be zeroed immediately after the parser succeeds, so an attacker has to write an
exploit that works without referencing data in the raw input.

Finally, Nail only supports error handling in a special debug configuration and will only print
error messages about the input to standard error. Besides complicating the parser, advanced error handling
invites programmers to attempt and fix malformed input, such as adding reasonable defaults for a
missing field. Such error-fixing not only introduces parser differential, but also might allow an
attacker to sneak inconsistent input past a parser.
% \paragraph{Intermediate representation.}

>>>>

\paragraph{Defense-in-depth.}

Security exploits often rely on raw inputs being present in memory~\cite{shotgun-parser}, for example to include
shell-code or crafted stack frames for ROP~\cite{phrack58:4-nergal} attacks in padding fields or the
application executing a controlled sequence of heap allocations and de-allocations to place
specific data at predictable addresses~\cite{jp-advanced, vudo-malloc}. Because the rest of the
application or even Nail's generated code may contain memory corruption bugs, Nail
carefully handles memory allocations as defense-in-depth to make exploiting such vulnerabilities harder.

When parsing input, Nail uses two separate memory arenas. These arenas allocate memory from the
system allocator in large, fixed-size blocks. Allocations are handled linearly and all data in the
arena is zeroed and freed at the same time. Nail uses one arena for data used only during parsing,
including dependent fields and temporary streams; this arena is released before the parser returns.
The other arena is used to allocate the internal data type returned and is freed by the application
once it is done processing an input.

Furthermore, the internal representation does not include any references to the input stream, which
can therefore be zeroed immediately after the parser succeeds, so an attacker has to write an
exploit that works without referencing data from the raw input.

Finally, Nail performs sophisticated error handling only in a special debug configuration and will print
error messages about the input only to \cc{stderr}. Besides complicating the parser, advanced error handling
invites programmers to attempt to fix malformed input, such as adding reasonable defaults for a
missing field. Such error-fixing not only introduces parser inconsistencies, but also might allow an
attacker to sneak inconsistent input past a parser.
% \paragraph{Intermediate representation.}
